Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-06-27 manuscript/main.tex
start improving unregularized estimation writing

\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}

For $k = 1, \ldots, K$, let us denote the system of $M$ basis functions used to approximate the $k$-th functional predictor as 
$
    \mathbf{b}^{(k)}(t_k)=(b^{(k)}_{1}(t_k), \ldots, b^{(k)}_{M}(t_k))^\top.
$
Then we can approximate the functional predictor observations as
\begin{equation*}
X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)=\mathbf{c}_i^{(k)T}\mathbf{b}^{(k)}(t_k), \;\; t_k \in \mathcal{T}_k, \;\; k=1, \ldots, K, \;\; i=1, \ldots, n,
\end{equation*}
where $\mathbf{c}_i^{(k)} = (c_{i1}^{(k)}, \ldots, c_{iM}^{(k)})^\top$ is a $M$-dimensional vector of basis coefficients corresponding to the $k$th functional predictor. This basis approximation can be expressed collectively across $n$ observations as: $\widetilde{\mathbf{X}^{(k)}}(t_k) = \widetilde{C}^{(k)} \mathbf{b}^{(k)}(t_k)$, where $\widetilde{\mathbf{X}}^{(k)}(t_k)=(X_1^{(k)}(t_k), \ldots, X_n^{(k)}(t_k))^\top \in \mathbb{R}^n$, and $\widetilde{C}^{(k)} = (\mathbf{c}_1^{(k)} \cdots \mathbf{c}_n^{(k)})^\top \in \mathbb{R}^{n \times M}$. This notation can be further extended to simultaneously express the basis expansions of $K$ functional predictors as 
\begin{equation}
    \widetilde{X}(\mathbf{t}) = \widetilde{C}B(\mathbf{t}), \quad t \in \mathcal{T},
    \label{eq: basis approximation of X}
\end{equation}
where $\widetilde{X}(\mathbf{t}) = [\widetilde{\mathbf{X}}^{(1)}(t_1), \cdots, \widetilde{\mathbf{X}}^{(K)}(t_K)] \in \mathbb{R}^{n \times K}$, $\widetilde{C} = [\widetilde{C}^{(1)}, \cdots, \widetilde{C}^{(K)}] \in \mathbb{R}^{n \times MK}$, and 
    $B(\mathbf{t}) = \mathrm{blkdiag}[\mathbf{b}^{(1)}(t_1), \cdots, \mathbf{b}^{(K)}(t_K)] \in \mathbb{R}^{MK \times K}$.
 The number of basis functions $M$ for each functional predictor can be set to a sufficiently large value (e.g., 15 or 20) to incorporate a rich variety of patterns that functional predictors may exhibit. Note that there is no need to carefully choose different $M$ values for different functional predictors, as an appropriate penalization scheme will be introduced in the later section to regularize each functional predictor to achieve a desired smoothness (see Section \ref{sec: Regularization of PLS Components}).

@@

We can also approximate the functional part of the PLS component in terms of the same basis functions: $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k) = \mathbf{a}^{(k)\top} \mathbf{b}^{(k)}(t_k)$ and $\psi(\mathbf{t}) = B(\mathbf{t})^\top \mathbf{a} \in \mathbb{R}^K$, where $\mathbf{a}^{(k)} = (a^{(k)}_{1}, \ldots, a^{(k)}_{M})^\top \in \mathbb{R}^M$, and $\mathbf{a}=(\mathbf{a}^{(1)\top}, \ldots, \mathbf{a}^{(K)\top})^\top \in \mathbb{R}^{MK}$, so that $\boldsymbol{\xi}[\mathbf{t}] = (\psi(\mathbf{t})^\top, \boldsymbol{\theta}^\top)^\top = (\mathbf{a}^\top B(\mathbf{t}), \boldsymbol{\theta}^\top)^\top \in \mathbb{R}^{K+p}$. We denote the outcome data as
\begin{equation}
\widetilde{\mathbf{y}}=(Y_1, \ldots, Y_n)^\top \in \mathbb{R}^n.
\end{equation}
%
For each $k = 1, \ldots, K$, let $J^{(k)} \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions associated with the $k$-th functional predictor, defined by
\begin{equation}\label{def:gram_basis}
J^{(k)} = \left[ \int_{\mathcal{T}k} b_m^{(k)}(t) \, b_n^{(k)}(t) , dt \right]_{m,n=1}^M.
\end{equation}
Using this definition, we construct the following block-diagonal matrices:
\begin{equation}\label{def:J_and_J_star}
J = \mathrm{blkdiag}(J^{(1)}, \cdots, J^{(K)}) \in \mathbb{R}^{MK \times MK}~\text{and}~J^\ast=\mathrm{blkdiag}(J, I_p) \in \mathbb{R}^{(MK+p) \times (MK+p)}.
\end{equation}

Our proposed strategy for computing PLS component, detailed in Section \ref{sec: Regularization of PLS Components}, incorporates regularization based on second-order derivatives. To this end, let $\ddot{J}^{(k)}$ denote the $M \times M$ Gram matrix formed by the second derivatives of the basis functions,  defined as
\begin{equation}\label{def:dodot_J_k}
\ddot{J}^{(k)} = \left[ \int_{\mathcal{T}_k} \ddot{b}^{(k)}_m(t) \ddot{b}^{(k)}_n(t) \, dt \right]_{m,n=1}^M.
\end{equation}
and define a block-diagonal matrix 
$ \ddot{J}^\ast \in \mathbb{R}^{(MK+p) \times (MK+p)}$ as
\begin{equation}\label{def:J_dotdot_ast}
    \ddot{J}^\ast=\mathrm{blkdiag}(\ddot{J}^{(1)}, \cdots, \ddot{J}^{(K)}, 0_{p \times p}).
\end{equation}
%
The finite-basis approximation allows us to represent the hybrid predictor observations using the coefficient matrix $\widetilde{C}^\ast$, enabling all associated computations to be carried out via matrix operations involving $\widetilde{C}^\ast$, $J^\ast$, and $\ddot{J}^\ast$.

\subsubsection{Smoothed estimation of PLS components}\label{sec: pls component computation intermediate}
Building on the basis function approximation from Section~\ref{sec: finite basis approximation}, we now introduce an alternative intermediate strategy for implementing the first step of the PLS algorithm. Specifically, the following proposition states a tractable generalized eigenvalue problem (i.e. maximization of a generalized Rayleigh quotient) that can be solved to directly obtain a hybrid PLS component $\widehat{\boldsymbol{\xi}}=(\widehat{\psi}, \widehat{\boldsymbol{\theta}})$, which has the same structure as the hybrid predictor $\mathbf{W}=(X,\mathbf{Z})$ of model \eqref{eq: Hybrid functional model}. 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}
At each $l$-th iteration of the first step of the PLS algorithm involving the residualized response and predictors $Y_i \equiv Y_i^{[l]}$ and $\mathbf{W}_i \equiv \mathbf{W}_i^{[l]}$, consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\jongmin{in the (1,1)th block, should $ \mathbf{y}$ be replaced by $\widetilde{\mathbf{y}} $? }
\begin{equation}\label{def:V_star}
and  $\boldsymbol{\xi}^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th PLS component, $\widehat{\boldsymbol{\xi}} \equiv \widehat{\boldsymbol{\xi}}_l$ can be obtained by solving
\begin{equation}\label{def:eigenproblem_smoothed}
       \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \; \frac{\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*}{\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*}, \quad \textit{or equivalently}, \quad \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \;  \boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^* \quad \textnormal{subject to}  \quad \boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1,
with respect to $\widehat{\boldsymbol{\xi}}^*=(\widehat{\mathbf{a}}^{\top}, \widehat{\boldsymbol{\theta}}^{\top})^\top$, and then setting $\widehat{\boldsymbol{\xi}}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^*[\mathbf{t}]^\top \widehat{\boldsymbol{\xi}}^*$. The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as $\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}$.
The proof of Proposition \ref{Proposition: generalized rayleigh quotient} is given in the Appendix \ref{section:proof:eigenproblem_derivation}.
Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.

>>>>

\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}
\paragraph{Basis functions.}
For $k = 1, \ldots, K$, let us denote the system of $M$ basis functions used to approximate the $k$-th functional predictor as 
\begin{equation*}
 \mathbf{b}^{(k)}(t_k)=(b^{(k)}_{1}(t_k), \ldots, b^{(k)}_{M}(t_k))^\top,~\text{where}~t_k \in \mathcal{T}_k.   
\end{equation*}
In our proposed hybrid PLS regression algorithm, this basis system is used to approximate both the predictors and the PLS components. It also serves as a regularization tool to borrow strength and prevent overfitting.

\paragraph{Basis approximation of the hybrid predictor.}
Then we can approximate the functional predictor observations as
\begin{equation*}
X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)=\mathbf{c}_i^{(k)T}\mathbf{b}^{(k)}(t_k), \;\; t_k \in \mathcal{T}_k, \;\; k=1, \ldots, K, \;\; i=1, \ldots, n,
\end{equation*}
where $\mathbf{c}_i^{(k)} = (c_{i1}^{(k)}, \ldots, c_{iM}^{(k)})^\top$ is a $M$-dimensional vector of basis coefficients corresponding to the $k$th functional predictor. This basis approximation can be expressed collectively across $n$ observations as: $\widetilde{\mathbf{X}^{(k)}}(t_k) = \widetilde{C}^{(k)} \mathbf{b}^{(k)}(t_k)$, where $\widetilde{\mathbf{X}}^{(k)}(t_k)=(X_1^{(k)}(t_k), \ldots, X_n^{(k)}(t_k))^\top \in \mathbb{R}^n$, and $\widetilde{C}^{(k)} = (\mathbf{c}_1^{(k)} \cdots \mathbf{c}_n^{(k)})^\top \in \mathbb{R}^{n \times M}$. This notation can be further extended to simultaneously express the basis expansions of $K$ functional predictors as 
\begin{equation}
    \widetilde{X}(\mathbf{t}) = \widetilde{C}B(\mathbf{t}), \quad t \in \mathcal{T},
    \label{eq: basis approximation of X}
\end{equation}
where $\widetilde{X}(\mathbf{t}) = [\widetilde{\mathbf{X}}^{(1)}(t_1), \cdots, \widetilde{\mathbf{X}}^{(K)}(t_K)] \in \mathbb{R}^{n \times K}$, $\widetilde{C} = [\widetilde{C}^{(1)}, \cdots, \widetilde{C}^{(K)}] \in \mathbb{R}^{n \times MK}$, and 
    $B(\mathbf{t}) = \mathrm{blkdiag}[\mathbf{b}^{(1)}(t_1), \cdots, \mathbf{b}^{(K)}(t_K)] \in \mathbb{R}^{MK \times K}$.
 The number of basis functions $M$ for each functional predictor can be set to a sufficiently large value (e.g., 15 or 20) to incorporate a rich variety of patterns that functional predictors may exhibit. Note that there is no need to carefully choose different $M$ values for different functional predictors, as an appropriate penalization scheme will be introduced in the later section to regularize each functional predictor to achieve a desired smoothness (see Section \ref{sec: Regularization of PLS Components}).

@@

\paragraph{Basis approximation of the PLS component.}
We can also approximate the functional part of the PLS component in terms of the same basis functions: $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k) = \mathbf{a}^{(k)\top} \mathbf{b}^{(k)}(t_k)$ and $\psi(\mathbf{t}) = B(\mathbf{t})^\top \mathbf{a} \in \mathbb{R}^K$, where $\mathbf{a}^{(k)} = (a^{(k)}_{1}, \ldots, a^{(k)}_{M})^\top \in \mathbb{R}^M$, and $\mathbf{a}=(\mathbf{a}^{(1)\top}, \ldots, \mathbf{a}^{(K)\top})^\top \in \mathbb{R}^{MK}$, so that $\boldsymbol{\xi}[\mathbf{t}] = (\psi(\mathbf{t})^\top, \boldsymbol{\theta}^\top)^\top = (\mathbf{a}^\top B(\mathbf{t}), \boldsymbol{\theta}^\top)^\top \in \mathbb{R}^{K+p}$. We denote the outcome data as
\begin{equation}
\widetilde{\mathbf{y}}=(Y_1, \ldots, Y_n)^\top \in \mathbb{R}^n.
\end{equation}
%
For each $k = 1, \ldots, K$, let $J^{(k)} \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions associated with the $k$-th functional predictor, defined by
\begin{equation}\label{def:gram_basis}
J^{(k)} = \left[ \int_{\mathcal{T}k} b_m^{(k)}(t) \, b_n^{(k)}(t) , dt \right]_{m,n=1}^M.
\end{equation}
Using this definition, we construct the following block-diagonal matrices:
\begin{equation}\label{def:J_and_J_star}
J = \mathrm{blkdiag}(J^{(1)}, \cdots, J^{(K)}) \in \mathbb{R}^{MK \times MK}~\text{and}~J^\ast=\mathrm{blkdiag}(J, I_p) \in \mathbb{R}^{(MK+p) \times (MK+p)}.
\end{equation}

\paragraph{Regularization based on second-order derivatives.}
Our proposed strategy for computing PLS component, detailed in Section \ref{sec: Regularization of PLS Components}, incorporates regularization based on second-order derivatives. To this end, let $\ddot{J}^{(k)}$ denote the $M \times M$ Gram matrix formed by the second derivatives of the basis functions,  defined as
\begin{equation}\label{def:dodot_J_k}
\ddot{J}^{(k)} = \left[ \int_{\mathcal{T}_k} \ddot{b}^{(k)}_m(t) \ddot{b}^{(k)}_n(t) \, dt \right]_{m,n=1}^M.
\end{equation}
and define a block-diagonal matrix 
$ \ddot{J}^\ast \in \mathbb{R}^{(MK+p) \times (MK+p)}$ as
\begin{equation}\label{def:J_dotdot_ast}
    \ddot{J}^\ast=\mathrm{blkdiag}(\ddot{J}^{(1)}, \cdots, \ddot{J}^{(K)}, 0_{p \times p}).
\end{equation}
%
The finite-basis approximation allows us to represent the hybrid predictor observations using the coefficient matrix $\widetilde{C}^\ast$, enabling all associated computations to be carried out via matrix operations involving $\widetilde{C}^\ast$, $J^\ast$, and $\ddot{J}^\ast$.

\subsubsection{Smoothed estimation of PLS components}\label{sec: pls component computation intermediate}
Building on the basis function approximation from Section~\ref{sec: finite basis approximation}, we now introduce an  intermediate strategy for implementing the first step of the PLS algorithm. Specifically, the following proposition states a tractable generalized eigenvalue problem (i.e. maximization of a generalized Rayleigh quotient) that can be solved to directly obtain a hybrid PLS component $\widehat{\boldsymbol{\xi}}=(\widehat{\psi}, \widehat{\boldsymbol{\theta}})$, which has the same structure as the hybrid predictor $\mathbf{W}=(X,\mathbf{Z})$ of model \eqref{eq: Hybrid functional model}.

At the $l$-th iteration, we denote, for simplicity, the residualized response and the hybrid predictors in basis approximation form as follows:
\begin{equation*}
    Y_i \equiv Y_i^{[l]} \quad \text{and}  \quad \mathbf{W}_i \equiv \mathbf{W}_i^{[l]},
    \widetilde{W}[\mathbf{t}]
\end{equation*}
respectively.
 consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}

\jongmin{in the (1,1)th block, should $ \mathbf{y}$ be replaced by $\widetilde{\mathbf{y}} $? }
\begin{equation}\label{def:V_star}
    V^\ast = n^{-2}\begin{bmatrix}
         J \widetilde{C}^\top \mathbf{y}\mathbf{y}^\top \widetilde{C} J & J \widetilde{C}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z} \\
        \widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C} J &   \widetilde{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{Z}
    \end{bmatrix} = n^{-2} J^\ast \widetilde{C}^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C}^{*} J^\ast
    \in \mathbb{R}^{(MK+p) \times (MK+p)}.
\end{equation}
and  $\boldsymbol{\xi}^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th PLS component, $\widehat{\boldsymbol{\xi}} \equiv \widehat{\boldsymbol{\xi}}_l$ can be obtained by solving
\begin{equation}\label{def:eigenproblem_smoothed}
       \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \; \frac{\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*}{\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*}, \quad \textit{or equivalently}, \quad \argmax \limits_{\boldsymbol{\xi}^* \in \mathbb{R}^{MK+p}} \;  \boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^* \quad \textnormal{subject to}  \quad \boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1,
\end{equation}
with respect to $\widehat{\boldsymbol{\xi}}^*=(\widehat{\mathbf{a}}^{\top}, \widehat{\boldsymbol{\theta}}^{\top})^\top$, and then setting $\widehat{\boldsymbol{\xi}}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^*[\mathbf{t}]^\top \widehat{\boldsymbol{\xi}}^*$. The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as $\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}$.
\end{proposition}
The proof of Proposition \ref{Proposition: generalized rayleigh quotient} is given in the Appendix \ref{section:proof:eigenproblem_derivation}.
Here, $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$ is expressed as $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ based on the aforementioned basis approximations. Thus, finding $\boldsymbol{\xi}^*$ that maximizes $\boldsymbol{\xi}^{*\top} V^\ast \boldsymbol{\xi}^*$ amounts to finding $\widehat{\boldsymbol{\xi}}$ that maximizes the sample squared covariance $\widehat{\textnormal{Cov}}^2(\langle \mathbf{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y)$, which is the first step of the PLS algorithm. The constraint $\boldsymbol{\xi}^{*\top} J^\ast \boldsymbol{\xi}^*=1$ scales the obtained PLS component to have a unit norm, that is, $\langle \widehat{\boldsymbol{\xi}}, \widehat{\boldsymbol{\xi}} \rangle_\mathcal{H}=1$.
